Simulations of fold number selection in cross-validation of linear and LASSO regression

### 2023-07-20 manuscript/manuscript.Rmd
“July20”

$k$-fold cross-validation is a popular method of error estimation for model selection in computational research. $n$ observations are divided into $k$ groups. In a first iteration, $i = k - 1$ groups are used as a training set. The remaining group is used as a test set to calculate estimated prediction error ($\mathrm{\widehat{PE}}$). This process is repeated $i$ times, with a different test set in each iteration. The average of $k$ estimates of $\mathrm{PE}$
$$\hat{\theta} = CV_{(k)} = \frac{1}{k}\sum_{i=1}^{k}\mathrm{\widehat{PE}}_i$$
is meant to estimate the true model error $\theta$, i.e., the error of the model tested on the population.

@@

More recently have been proposed ways to quantify the variance reduction achieved by cross-validation when the true prediction error ($\mathrm{PE}$) is not known, e.g., as mean-square stability (@Kale2011), where
$$\underset{S,z,z'}{\mathrm{E}}[(\ell_{\mathcal A(S)}(z) - \ell_{\mathcal A(S^{i,z'})}(z))^2] \le \beta,$$
i.e., the expectation of the squared difference in loss $\ell$ of an algorithm $\mathcal A$ over a dataset $S$ and over a dataset $S'$ such that one element $z' \sim Z$ has been added to $S$ is less than $\beta$, which depends on the algorithm; or as loss stability (@Kumar2013), where
$$\underset{S,z,z'}{\mathrm{E}}[(\ell'_{\mathcal A(S)}(z) - \ell'_{\mathcal A(S^{i,z'})}(z))^2] \le \gamma,$$
$\ell'_{\mathcal A(S)}(z) = \ell_{\mathcal A(S)}(z) - \bar{\ell}_{\mathcal A(S)}(z)$, and $\gamma$ depends on the algorithm. However, it has also been demonstrated that, due to overlap between training and test sets in cross-validation, there is no universal (i.e., valid under all distributions) unbiased estimator of the variance of $k$-fold cross-validation (@Bengio2004).

@@

Consider a population of size N = 500,000 with five of $P = 100$ features $X_1...X_{100} \sim N(0, 1)$ a linear combination of feature $Y = \beta_0 + \beta_1X_1 + \cdots + \beta_5X_5 + \epsilon$
where $\beta_1 = \cdots = \beta_5 = 1$ and $\epsilon \sim N(0, 10)$. The true model of $Y$ is $f \in F$, a set of competing models. $f = \mathrm{E}(Y) = \beta_0 + \beta_1X_1 + \cdots + \beta_5X_5$
and its mean squared error ($\mathrm{MSE}$) is
$$\theta = \mathrm{MSE}[\mathrm{E}(Y)] = \frac{\sum_{i=1}^{N}[Y_i - \mathrm{E}(Y_i)]^2}{N}.$$

@@

This is because in practice the model with the lowest $\mathrm{\widehat{PE}}$ is more likely to be selected. We refer to this as lowest-error model selection. As our simulation results show, however, it is possible for $k$-fold cross-validation to result in calculations of $\mathrm{MSE}(\hat{\theta})$ such that a competing model $\hat{f}$ has a lower $\mathrm{\widehat{PE}}$ than the true model $f$. This false model may have the lowest $\mathrm{\widehat{PE}}$ but its generalization error will be higher in the long-run (i.e., when tested on a large part of the population) than the generalization error of the true model. Thus, it is preferable for the value of $k$ selected to result in $f$ having the lowest $\mathrm{\widehat{PE}}$.

@@

To estimate $k_\mathrm{optimal}*$ for each $n$, we first fit to the dataset $(k,A)$ for each sample size $n$ a non-linear least squares model of the form
$$\hat{A}(k) = Bk + C + Dk^{-1} + Ek^{-2}$$
with constants $B$ to $E$ specific to each $n$ and constants $B$ and $E$ set to zero for linear regression. We then draw a line $L$ through $(k_1, \hat{A}_1)$ and $(k_n, \hat{A}_n)$ and maximize the perpendicular distance $d$ of each point on $\hat{A}$ from $L$, so that
$$k_\mathrm{optimal}^* = {\arg\max}_k{d[\hat{A}(k)-L]}.$$

@@

```{r echo=FALSE, fig.height = 8, fig.cap = "Count of true linear regression model selection by the lowest-error criterion vs. fold number for various sample sizes."}
figure7
```

```{r echo=FALSE, fig.cap = "Count of true LASSO regression model selection by the lowest-error criterion vs. fold number for various sample sizes."}
figure8
```

>>>>

$k$-fold cross-validation is a popular method of error estimation for model selection in computational research. $n$ observations are divided into $k$ groups. In a first iteration, $i = k - 1$ groups are used as a training set. The remaining group is used as a test set to calculate estimated prediction error ($\mathrm{\widehat{PE}}$). This process is repeated $i$ times, with a different test set in each iteration. The average of $k$ estimates of $\mathrm{PE}$
$$\hat{\theta} = CV_{(k)} = \frac{1}{k}\sum_{i=1}^{k}\mathrm{\widehat{PE}}_i$$
is meant to estimate the true model error $\theta$, i.e., the error of the model tested on the population (@Bates2022).

@@

More recently have been proposed ways to quantify the variance reduction achieved by cross-validation when the true prediction error ($\mathrm{PE}$) is not known, e.g., as mean-square stability (@Kale2011) or as loss stability (@Kumar2013). However, it has also been demonstrated that, due to overlap between training and test sets in cross-validation, there is no universal (i.e., valid under all distributions) unbiased estimator of the variance of $k$-fold cross-validation (@Bengio2004).

@@

In machine learning tasks, selection of $k$ for $k$-fold cross validation is largely an arbitrary decision between, e.g., 5 and 10. With the results of the following simulation, we make a less arbitrary recommendation of $k$ with the aim of improving model selection accuracy.

Consider a population of size N = 500,000 with five of $P = 100$ features $X_1...X_{100} \sim N(0, 1)$ a linear combination of feature $Y = \beta_0 + \beta_1X_1 + \cdots + \beta_5X_5 + \epsilon$
where $\beta_1 = \cdots = \beta_5 = 1$ and $\epsilon \sim N(0, 10)$. The true model of $Y$ is $f \in F$, a set of competing models. $f = \mathrm{E}(Y) = \beta_0 + \beta_1X_1 + \cdots + \beta_5X_5$
and its mean squared error ($\mathrm{MSE}$) is
$$\theta = \mathrm{MSE}(f) = \frac{\sum_{i=1}^{N}[Y_i - \mathrm{E}(Y_i)]^2}{N}.$$

@@

This is because in practice the model with the lowest $\mathrm{\widehat{PE}}$ is more likely to be selected. We refer to this as lowest-error model selection. As our simulation results show, however, it is possible for $k$-fold cross-validation to result in calculations of $\mathrm{MSE}(\hat{\theta})$ such that a competing model $\hat{f} \in F$ has a lower $\mathrm{\widehat{PE}}$ than the true model $f$. This false model may have the lowest $\mathrm{\widehat{PE}}$ but its generalization error will be higher in the long-run (i.e., when tested on a large part of the population) than the generalization error of the true model. Thus, it is preferable for the value of $k$ selected to result in $f$ having the lowest $\mathrm{\widehat{PE}}$.

@@

To estimate $k_\mathrm{optimal}*$ for each $n$, we first fit to the dataset $(k,A)$ for each sample size $n$ a model of the form
$$\hat{A}(k) = Bk + C + \frac{D}{k}$$
with constants $B$ to $D$ specific to each $n$. Certain more complicated models (e.g., with higher-order terms) may provide better approximation of $A$ but make optimization by the following scheme non-trivial. We draw a line $L$ through $(k_1, \hat{A}_1)$ and $(k_n, \hat{A}_n)$ and maximize the perpendicular distance $d$ of each point on $\hat{A}$ from $L$, so that
\begin{align*}
k^*_{\mathrm{optimal}} &= {\arg\max}_k{d[\hat{A}(k),L]} \\
                       &= {\arg\max}_k{d[Bk+C+\frac{D}{k},y-(Bn+C+\frac{D}{n})=m(x-n)]} \\
                       &= {\arg\max}_k{d[Bk+C+\frac{D}{k},y-(Bn+C+\frac{D}{n})=\frac{(Bn+C+\frac{D}{n})-(2B+C+\frac{D}{2})}{n-2}(x-n)]} \\
                       &= {\arg\max}_k{d[Bk+C+\frac{D}{k},y-(Bn+C+\frac{D}{n})=(B-\frac{D}{2n})(x-n)]} \\
                       &= {\arg\max}_k{d[Bk+C+\frac{D}{k},(B-\frac{D}{2n})x-y+C+\frac{D}{2}+\frac{D}{n}=0]}. \\
\end{align*}
It is known that the distance $d$ between a point $(k, \hat{A}(k))$ and a line of the form
$$ax + by + c = 0$$
is
\begin{align*}
d = \frac{|ak + b\hat{A}(k) + c|}{\sqrt{a^2+b^2}}
\end{align*}
so
\begin{align*}
k^*_{\mathrm{optimal}} &= {\arg\max}_k{d[Bk+C+\frac{D}{k},(B-\frac{D}{2n})x-y+C+\frac{D}{2}+\frac{D}{n}=0]} \\
                       &= {\arg\max}_k{\frac{|(B-\frac{D}{2n})k-(Bk+C+\frac{D}{k})+(C+\frac{D}{2}+\frac{D}{n})|}{\sqrt{(B-\frac{D}{2n})^2+(-1)^2}}}.
\end{align*}
Maximizing $d$, knowing that $n$ and $k$ are always positive, we find that
\begin{align*}
\frac{d}{dk}d[\hat{A}(k),L] &= \frac{d}{dk}\left(\frac{|(B-\frac{D}{2n})k-(Bk+C+\frac{D}{k})+(C+\frac{D}{2}+\frac{D}{n})|}{\sqrt{(B-\frac{D}{2n})^2+(-1)^2}}\right) \\
                            &=\frac{|n||k|(2nD-Dk^2)(-Dk^2-2nD+nDk+2Dk)}{nk^3|-Dk^2-2nD+nDk+2Dk|\sqrt{D^2+4n^2}} = 0 \\
                            &\Rightarrow k = 0, 2, \sqrt{2n}, n.
\end{align*}
Evaluating $d$ at these four values of $k$, we find that
\begin{align*}
d(0) &= \mathrm{DNE} \\
d(2) &= 0 \\
d(\sqrt{2n}) &=  \frac{|nD+2D-\sqrt{2n}D-D|}{\sqrt{(B-\frac{D}{2n})^2+(-1)^2}} \\
d(n) &= 0.
\end{align*}
Since $d(\sqrt{2n}) > 0$,
$$k^*_{\mathrm{optimal}} = \sqrt{2n}$$
which fits the relationship of $k^*_{\mathrm{optimal}}$ vs. $n$ (Figure 9).

@@

```{r echo=FALSE, fig.height = 8, fig.cap = "Count of true linear regression model selection by the lowest-error criterion vs. fold number for various sample sizes. Red line indicates optimal value of k."}
figure7
```

```{r echo=FALSE, fig.cap = "Count of true LASSO regression model selection by the lowest-error criterion vs. fold number for various sample sizes. Red line indicates optimal value of k."}
figure8
```

### 2023-07-20 manuscript/manuscript.tex
“July20”

\(k\)-fold cross-validation is a popular method of error estimation for
model selection in computational research. \(n\) observations are
divided into \(k\) groups. In a first iteration, \(i = k - 1\) groups
are used as a training set. The remaining group is used as a test set to
calculate estimated prediction error (\(\mathrm{\widehat{PE}}\)). This
process is repeated \(i\) times, with a different test set in each
iteration. The average of \(k\) estimates of \(\mathrm{PE}\)
\[\hat{\theta} = CV_{(k)} = \frac{1}{k}\sum_{i=1}^{k}\mathrm{\widehat{PE}}_i\]
is meant to estimate the true model error \(\theta\), i.e., the error of
the model tested on the population.

@@

More recently have been proposed ways to quantify the variance reduction
achieved by cross-validation when the true prediction error
(\(\mathrm{PE}\)) is not known, e.g., as mean-square stability (Kale,
Kumar, and Vassilvitskii (\protect\hyperlink{ref-Kale2011}{2011})),
where
\[\underset{S,z,z'}{\mathrm{E}}[(\ell_{\mathcal A(S)}(z) - \ell_{\mathcal A(S^{i,z'})}(z))^2] \le \beta,\]
i.e., the expectation of the squared difference in loss \(\ell\) of an
algorithm \(\mathcal A\) over a dataset \(S\) and over a dataset \(S'\)
such that one element \(z' \sim Z\) has been added to \(S\) is less than
\(\beta\), which depends on the algorithm; or as loss stability (Kumar
et al. (\protect\hyperlink{ref-Kumar2013}{2013})), where
\[\underset{S,z,z'}{\mathrm{E}}[(\ell'_{\mathcal A(S)}(z) - \ell'_{\mathcal A(S^{i,z'})}(z))^2] \le \gamma,\]
\(\ell'_{\mathcal A(S)}(z) = \ell_{\mathcal A(S)}(z) - \bar{\ell}_{\mathcal A(S)}(z)\),
and \(\gamma\) depends on the algorithm. However, it has also been
demonstrated that, due to overlap between training and test sets in
cross-validation, there is no universal (i.e., valid under all
distributions) unbiased estimator of the variance of \(k\)-fold
cross-validation (Bengio and Grandvalet
(\protect\hyperlink{ref-Bengio2004}{2004})).

@@

Consider a population of size N = 500,000 with five of \(P = 100\)
features \(X_1...X_{100} \sim N(0, 1)\) a linear combination of feature
\(Y = \beta_0 + \beta_1X_1 + \cdots + \beta_5X_5 + \epsilon\) where
\(\beta_1 = \cdots = \beta_5 = 1\) and \(\epsilon \sim N(0, 10)\). The
true model of \(Y\) is \(f \in F\), a set of competing models.
\(f = \mathrm{E}(Y) = \beta_0 + \beta_1X_1 + \cdots + \beta_5X_5\) and
its mean squared error (\(\mathrm{MSE}\)) is
\[\theta = \mathrm{MSE}[\mathrm{E}(Y)] = \frac{\sum_{i=1}^{N}[Y_i - \mathrm{E}(Y_i)]^2}{N}.\]

@@

This is because in practice the model with the lowest
\(\mathrm{\widehat{PE}}\) is more likely to be selected. We refer to
this as lowest-error model selection. As our simulation results show,
however, it is possible for \(k\)-fold cross-validation to result in
calculations of \(\mathrm{MSE}(\hat{\theta})\) such that a competing
model \(\hat{f}\) has a lower \(\mathrm{\widehat{PE}}\) than the true
model \(f\). This false model may have the lowest
\(\mathrm{\widehat{PE}}\) but its generalization error will be higher in
the long-run (i.e., when tested on a large part of the population) than
the generalization error of the true model. Thus, it is preferable for
the value of \(k\) selected to result in \(f\) having the lowest
\(\mathrm{\widehat{PE}}\).

@@

To estimate \(k_\mathrm{optimal}*\) for each \(n\), we first fit to the
dataset \((k,A)\) for each sample size \(n\) a non-linear least squares
model of the form \[\hat{A}(k) = Bk + C + Dk^{-1} + Ek^{-2}\] with
constants \(B\) to \(E\) specific to each \(n\) and constants \(B\) and
\(E\) set to zero for linear regression. We then draw a line \(L\)
through \((k_1, \hat{A}_1)\) and \((k_n, \hat{A}_n)\) and maximize the
perpendicular distance \(d\) of each point on \(\hat{A}\) from \(L\), so
that \[k_\mathrm{optimal}^* = {\arg\max}_k{d[\hat{A}(k)-L]}.\]

@@

\caption{Count of true linear regression model selection by the lowest-error criterion vs. fold number for various sample sizes.}\label{fig:unnamed-chunk-7}
\end{figure}

@@

\caption{Count of true LASSO regression model selection by the lowest-error criterion vs. fold number for various sample sizes.}\label{fig:unnamed-chunk-8}
\end{figure}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-Bengio2004}{}}%
Bengio, Yoshua, and Yves Grandvalet. 2004. {``No Unbiased Estimator of
the Variance of k-Fold Cross-Validation.''} \emph{Journal of Machine
Learning Research} 5 (June): 1089--1105.

>>>>

\(k\)-fold cross-validation is a popular method of error estimation for
model selection in computational research. \(n\) observations are
divided into \(k\) groups. In a first iteration, \(i = k - 1\) groups
are used as a training set. The remaining group is used as a test set to
calculate estimated prediction error (\(\mathrm{\widehat{PE}}\)). This
process is repeated \(i\) times, with a different test set in each
iteration. The average of \(k\) estimates of \(\mathrm{PE}\)
\[\hat{\theta} = CV_{(k)} = \frac{1}{k}\sum_{i=1}^{k}\mathrm{\widehat{PE}}_i\]
is meant to estimate the true model error \(\theta\), i.e., the error of
the model tested on the population (Bates, Hastie, and Tibshirani
(\protect\hyperlink{ref-Bates2022}{2022})).

@@

More recently have been proposed ways to quantify the variance reduction
achieved by cross-validation when the true prediction error
(\(\mathrm{PE}\)) is not known, e.g., as mean-square stability (Kale,
Kumar, and Vassilvitskii (\protect\hyperlink{ref-Kale2011}{2011})) or as
loss stability (Kumar et al. (\protect\hyperlink{ref-Kumar2013}{2013})).
However, it has also been demonstrated that, due to overlap between
training and test sets in cross-validation, there is no universal (i.e.,
valid under all distributions) unbiased estimator of the variance of
\(k\)-fold cross-validation (Bengio and Grandvalet
(\protect\hyperlink{ref-Bengio2004}{2004})).

@@

In machine learning tasks, selection of \(k\) for \(k\)-fold cross
validation is largely an arbitrary decision between, e.g., 5 and 10.
With the results of the following simulation, we make a less arbitrary
recommendation of \(k\) with the aim of improving model selection
accuracy.

Consider a population of size N = 500,000 with five of \(P = 100\)
features \(X_1...X_{100} \sim N(0, 1)\) a linear combination of feature
\(Y = \beta_0 + \beta_1X_1 + \cdots + \beta_5X_5 + \epsilon\) where
\(\beta_1 = \cdots = \beta_5 = 1\) and \(\epsilon \sim N(0, 10)\). The
true model of \(Y\) is \(f \in F\), a set of competing models.
\(f = \mathrm{E}(Y) = \beta_0 + \beta_1X_1 + \cdots + \beta_5X_5\) and
its mean squared error (\(\mathrm{MSE}\)) is
\[\theta = \mathrm{MSE}(f) = \frac{\sum_{i=1}^{N}[Y_i - \mathrm{E}(Y_i)]^2}{N}.\]

@@

This is because in practice the model with the lowest
\(\mathrm{\widehat{PE}}\) is more likely to be selected. We refer to
this as lowest-error model selection. As our simulation results show,
however, it is possible for \(k\)-fold cross-validation to result in
calculations of \(\mathrm{MSE}(\hat{\theta})\) such that a competing
model \(\hat{f} \in F\) has a lower \(\mathrm{\widehat{PE}}\) than the
true model \(f\). This false model may have the lowest
\(\mathrm{\widehat{PE}}\) but its generalization error will be higher in
the long-run (i.e., when tested on a large part of the population) than
the generalization error of the true model. Thus, it is preferable for
the value of \(k\) selected to result in \(f\) having the lowest
\(\mathrm{\widehat{PE}}\).

@@

To estimate \(k_\mathrm{optimal}*\) for each \(n\), we first fit to the
dataset \((k,A)\) for each sample size \(n\) a model of the form
\[\hat{A}(k) = Bk + C + \frac{D}{k}\] with constants \(B\) to \(D\)
specific to each \(n\). Certain more complicated models (e.g., with
higher-order terms) may provide better approximation of \(A\) but make
optimization by the following scheme non-trivial. We draw a line \(L\)
through \((k_1, \hat{A}_1)\) and \((k_n, \hat{A}_n)\) and maximize the
perpendicular distance \(d\) of each point on \(\hat{A}\) from \(L\), so
that \begin{align*}
k^*_{\mathrm{optimal}} &= {\arg\max}_k{d[\hat{A}(k),L]} \\
                       &= {\arg\max}_k{d[Bk+C+\frac{D}{k},y-(Bn+C+\frac{D}{n})=m(x-n)]} \\
                       &= {\arg\max}_k{d[Bk+C+\frac{D}{k},y-(Bn+C+\frac{D}{n})=\frac{(Bn+C+\frac{D}{n})-(2B+C+\frac{D}{2})}{n-2}(x-n)]} \\
                       &= {\arg\max}_k{d[Bk+C+\frac{D}{k},y-(Bn+C+\frac{D}{n})=(B-\frac{D}{2n})(x-n)]} \\
                       &= {\arg\max}_k{d[Bk+C+\frac{D}{k},(B-\frac{D}{2n})x-y+C+\frac{D}{2}+\frac{D}{n}=0]}. \\
\end{align*} It is known that the distance \(d\) between a point
\((k, \hat{A}(k))\) and a line of the form \[ax + by + c = 0\] is
\begin{align*}
d = \frac{|ak + b\hat{A}(k) + c|}{\sqrt{a^2+b^2}}
\end{align*} so \begin{align*}
k^*_{\mathrm{optimal}} &= {\arg\max}_k{d[Bk+C+\frac{D}{k},(B-\frac{D}{2n})x-y+C+\frac{D}{2}+\frac{D}{n}=0]} \\
                       &= {\arg\max}_k{\frac{|(B-\frac{D}{2n})k-(Bk+C+\frac{D}{k})+(C+\frac{D}{2}+\frac{D}{n})|}{\sqrt{(B-\frac{D}{2n})^2+(-1)^2}}}.
\end{align*} Maximizing \(d\), knowing that \(n\) and \(k\) are always
positive, we find that \begin{align*}
\frac{d}{dk}d[\hat{A}(k),L] &= \frac{d}{dk}\left(\frac{|(B-\frac{D}{2n})k-(Bk+C+\frac{D}{k})+(C+\frac{D}{2}+\frac{D}{n})|}{\sqrt{(B-\frac{D}{2n})^2+(-1)^2}}\right) \\
                            &=\frac{|n||k|(2nD-Dk^2)(-Dk^2-2nD+nDk+2Dk)}{nk^3|-Dk^2-2nD+nDk+2Dk|\sqrt{D^2+4n^2}} = 0 \\
                            &\Rightarrow k = 0, 2, \sqrt{2n}, n.
\end{align*} Evaluating \(d\) at these four values of \(k\), we find
that \begin{align*}
d(0) &= \mathrm{DNE} \\
d(2) &= 0 \\
d(\sqrt{2n}) &=  \frac{|nD+2D-\sqrt{2n}D-D|}{\sqrt{(B-\frac{D}{2n})^2+(-1)^2}} \\
d(n) &= 0.
\end{align*} Since \(d(\sqrt{2n}) > 0\),
\[k^*_{\mathrm{optimal}} = \sqrt{2n}\] which fits the relationship of
\(k^*_{\mathrm{optimal}}\) vs.~\(n\) (Figure 9).

@@

\caption{Count of true linear regression model selection by the lowest-error criterion vs. fold number for various sample sizes. Red line indicates optimal value of k.}\label{fig:unnamed-chunk-7}
\end{figure}

@@

\caption{Count of true LASSO regression model selection by the lowest-error criterion vs. fold number for various sample sizes. Red line indicates optimal value of k.}\label{fig:unnamed-chunk-8}
\end{figure}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-Bates2022}{}}%
Bates, Stephen, Trevor Hastie, and Robert Tibshirani. 2022.
{``Cross-Validation: What Does It Estimate and How Well Does It Do
It?''} \url{https://arxiv.org/abs/2104.00673}.

\leavevmode\vadjust pre{\hypertarget{ref-Bengio2004}{}}%
Bengio, Yoshua, and Yves Grandvalet. 2004. {``No Unbiased Estimator of
the Variance of k-Fold Cross-Validation.''} \emph{Journal of Machine
Learning Research} 5 (June): 1089--1105.
